Preserving Reciprocal Consistency in Distributed Graph Databases

### 2020-02-17 paper/main.tex
add algorithm templates
making results

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A solution for maintaining reciprocal consistency in distributed graph databases is presented.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{reciprocal consistency, graph databases, collision detection}

%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{Seattle Mariners at Spring Training, 2010.}
%   \Description{Enjoying the baseball game from the third-base
%   seats. Ichiro Suzuki preparing to bat.}
%   \label{fig:teaser}
% \end{teaserfigure}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
In recent years there has been an increased interest in and use of graph technologies[REF]. Application areas are wide reaching from healthcare to more traditional use cases such as social networks [REF]. Graph databases model as a \textit{labeled property graph}, vertices represent entities and edges represent the relationship between entities; properties can be stored on both vertices and edges [REF].

In practice graphs can be extremely large [REF], exceeding the storage capacity of a single node database. An approach to storing large graphs is to partition graph data over several machines in a cluster [REF]. Moving from a single-node database to one partitioned and distributed across servers introduces several interesting problems regarding data consistency. Graphs place a high demand on the underlying database and maintaining structural integrity across partitions - called \textit{reciprocal consistency} - in a distributed graph database is non-trivial.

Recent work has investigated how violation of reciprocal consistency can lead to semantic corruption at alarmingly rates [REF]. In order to prevent this form of inconsistency, one could implement a heavyweight coordination protocol across partitions providing Serializability or Snapshot Isolation, which have associated performance costs [REF]. The approach taken in this paper is to build on earlier work and develop a protocol that provides weaker consistency guarantees whilst still preserving reciprocal consistency.

@@

In the labelled property graph data model edges have direction, there is a \textit{source} and a \textit{destination} vertex. However, for query performance, edge information is stored with \textbf{both} the source and destination vertices. This facilitates bi-directional edge traversal. Graphs are often implemented as an collection of adjacency lists where a list contains neighbors of a given vertex [REF].

Consider, for example, the statement that Tolkien \textit{wrote} The Hobbit. It is expressed using vertices A and B, for Tolkien and The Hobbit respectively, and an edge \textit{wrote} running from A (source) to B (destination). Adjacency lists of both A and B record information about that edge and this information is mutually reciprocal (or inverse) of each other: A's list will indicate `A \textit{wrote} B' while B's will have `B \textit{written} by A'. Thus, a query `list all titles by the author who wrote The Hobbit' can be answered by landing at B and then traversing to A, even though AB is a directed edge at model level abstraction.

The reciprocal nature of information at the ends of an edge not only facilitates bidirectional traversal but also ensures operational correctness. Without it, the queries `who wrote The Hobbit?' and `what all did Tolkien write?' could yield in different results. So, an edge is said to be \textit{reciprocally consistent}, if the adjacency lists of the vertices it connects have mutually reciprocal information.

@@

A common approach to partitioning a graph is to use a balanced edge-cut. Vertices are evenly assigned to partitions in a manner that minimises the number of edges that span partitions - referred to as \textbf{distributed edges}. Regardless of the partitioning strategy used a non-negligible number of edges are distributed edges.

When concurrent queries update an edge, maintaining reciprocal consistency gets harder if the edge is distributed, i.e., its vertices are stored in distinct servers of a distributed database system. Our earlier work [REF] established that when edges are permitted to become reciprocally inconsistent, independent queries reading inconsistent information from different ends of such edges, do corrupt the database through their subsequent writes, and that a significant degree of corruption can result within durations that are worryingly small compared to database lifetime.

Of course, reciprocal consistency can be maintained as a part of enforcing any of several known strong isolation guarantees (e.g., snapshot isolation) but that incurs a performance cost if users are interested only in ensuring reciprocal consistency to preserve the integrity of a distributed database. This paper addresses that
challenge. It proposes a suite of simple, coordination-free protocols and compares their performance through simulations.

@@

Both protocols described below leverage the same fact: at the logical level, updating a distributed edge is a single operation. However, it consists of two sub-operations, it must update information at the source and destination vertices an edge connects. Note, the order in which the sub-operations take place is not constrained. Updating the source then the destination is equally as likely as updating the destination then the source.

\subsection{Probabilistic Protocol}

In the probabilistic approach, an edge update is allowed if an earlier, temporary update is sufficiently early by D as per the local clock time and D is the bound on the server communication delays with probability (1-e). (Note that a more pessimistic estimation of D results in a smaller e) In this approach, the smaller the value chosen for D, the more is the extent of contentions permitted, hence the smaller is the number of expected aborts and, on the other hand, the larger is the risk of reciprocal consistency not being ensured and the database being corrupted. These are examined through simulations.

\subsection{Deterministic Protocol}

In the deterministic approach, queries updating distributed edges are expected to carry with them some state information as they traverse from one end server of a distributed edge to the other. Other than this overhead, there is no requirement for estimating bound on traversal delays. Our simulations will compare the number of aborts incurred in both approaches.

@@

@@

% % The bibliography is included in your source document with these two
% % commands, placed just before the \verb|\end{document}| command:
% % \begin{verbatim}
% %   \bibliographystyle{ACM-Reference-Format}
% %   \bibliography{bibfile}
% % \end{verbatim}
% % where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
% % suffix, of the \BibTeX\ file.

@@

% % \section{Appendices}

% % If your work needs an appendix, add it before the
% % ``\verb|\end{document}|'' command at the conclusion of your source
% % document.

% % Start the appendix with the ``\verb|appendix|'' command:
% % \begin{verbatim}
% %   \appendix
% % \end{verbatim}
% % and note that in the appendix, sections are lettered, not
% % numbered. This document has two appendices, demonstrating the section
% % and subsection identification method.

% % \section{SIGCHI Extended Abstracts}

% % The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
% % not in Word) produces a landscape-orientation formatted article, with
% % a wide left margin. Three environments are available for use with the
% % ``\verb|sigchi-a|'' template style, and produce formatted output in
% % the margin:
% % \begin{itemize}
% % \item {\verb|sidebar|}:  Place formatted text in the margin.
% % \item {\verb|marginfigure|}: Place a figure in the margin.
% % \item {\verb|margintable|}: Place a table in the margin.
% % \end{itemize}

% % %%
% % %% The acknowledgments section is defined using the "acks" environment
% % %% (and NOT an unnumbered section). This ensures the proper
% % %% identification of the section in the article metadata, and the
% % %% consistent spelling of the heading.
% % \begin{acks}
% % To Robert, for the bagels and explaining CMYK and color spaces.
% % \end{acks}

% % %%
% % %% The next two lines define the bibliography style to be used, and
% % %% the bibliography file.
% % \bibliographystyle{ACM-Reference-Format}
% % \bibliography{sample-base}

@@

% % \section{Research Methods}

% % \subsection{Part One}

% % Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
% % malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
% % sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
% % vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
% % lacinia dolor. Integer ultricies commodo sem nec semper.

% % \subsection{Part Two}

% % Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
% % ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
% % ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
% % eros. Vivamus non purus placerat, scelerisque diam eu, cursus
% % ante. Etiam aliquam tortor auctor efficitur mattis.

% % \section{Online Resources}

% % Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
% % pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
% % enim maximus. Vestibulum gravida massa ut felis suscipit
% % congue. Quisque mattis elit a risus ultrices commodo venenatis eget
% % dui. Etiam sagittis eleifend elementum.

% % Nam interdum magna at lectus dignissim, ac dignissim lorem
% % rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
% % massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

>>>>

\documentclass[sigplan,screen]{acmart}

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\usepackage[ruled,vlined]{algorithm2e}

@@

\begin{abstract}

  In this paper we formalize the notion of \textit{reciprocal consistency}, a important feature specific to graph databases. If reciprocal consistency is not enforced during transaction processing, a scale-free distributed graph database can become semantically corrupt within a short time period relative to database lifetime. Reciprocal consistency can be maintained as a part of enforcing any of several known strong isolation guarantees (e.g., snapshot isolation) incurring well established performance costs. However, in practice systems are often deployed with weaker consistency, this paper describes two protocols that provide only reciprocal consistency semantics. Catering for users that are interested only in ensuring reciprocal consistency to preserve the integrity of a distributed graph database. Both protocols presented are compared and contrasted using metrics such as number of aborts per second.

\end{abstract}

% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

@@

\keywords{Graph Databases, Reciprocal Consistency}

\maketitle

\section{Introduction}

Recent years have seen a proliferation in the use of graph technologies \cite{Besta2019}. Application areas are wide reaching from healthcare, social networks and fraud detection \cite{Eifrem2016}. Graph databases model data as a \textit{labeled property graph}, vertices represent entities and edges represent the relationship between entities; properties can be stored on both vertices and edges \cite{Robinson2015}.

In practice graphs can be extremely large \cite{Sahu2017}, exceeding the storage capacity of a single-node database. An approach to storing large graphs is to partition graph data over several machines in a cluster. Moving from a single-node database to one partitioned and distributed across servers introduces several interesting problems regarding data consistency. Graphs place a high demand on the underlying database and maintaining structural integrity across partitions - called \textit{reciprocal consistency} - in a distributed graph database is non-trivial.

In order to preserve reciprocal consistency, one could implement a heavyweight coordination protocol across partitions providing Serializability or Snapshot Isolation, which have associated performance costs. To avoid these costs database managements systems (DBMSs) provide several weak isolation models \cite{Berenson1995}. In addition, there are a number of systems that eschew transactional guarantees altogether, typically offering eventual consistency \cite{Bailis2013}. Recent work \cite{Ezhilchelvan2018} investigated how violation of reciprocal consistency can occur under eventual consistency, leading to semantic corruption at alarmingly rates. The contribution of this paper is the description of two protocols that provide only reciprocal consistency, a consistency model between the known extremes that provides suitable performance. The performance of each protocol is measured using metrics such as aborts per second.

@@

In the labeled property graph data model edges have direction, there is a \textit{source} and a \textit{destination} vertex. Edge information is stored with \textbf{both} the source and destination vertices. This facilitates bi-directional edge traversal and allows for better query performance.

A common approach to storing graphs (arising from JanusGraph \cite{janusgraph} and TitanDB [REF]) is for records to represent vertices containing both data values and an adjacency list containing edge ``pointers'' to other vertices. In this representation an edge has ``reciprocal'' entries in the adjacency lists of the vertices the edge connects Figure \ref{adj-list}. A query reading either the source or destination vertices should be able to reify the edge correctly, returning consistent results. When the adjacency lists for vertices are mutually compatible like this, we say they are \emph{reciprocally consistent}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{./images/janusgraph-adj-list}
  \caption{Graph representation [REF].}
  \Description{Graph representation.}
  \label{adj-list}
\end{figure}

Consider, for example, the statement that Tolkien \textit{wrote} The Hobbit. It is expressed using vertices A and B, for Tolkien and The Hobbit respectively, and an edge \textit{wrote} running from A (source) to B (destination). Adjacency lists of both A and B record information about that edge and this information is mutually reciprocal (or inverse) of each other: A's list will indicate `A \textit{wrote} B' while B's will have `B \textit{written} by A'. Thus, a query `list all titles by the author who wrote The Hobbit' can be answered by landing at B and then traversing to A; even though the edge \texttt{(A \{name:`Tolkien'\})-[:WROTE]->(B \{title:`The Hobbit'\})} is a directed edge at model level abstraction.

@@

A distributed graph database employs a shared-nothing architecture, partitioning a graph between a number of loosely cooperating servers. Graph partitioning is non-trivial and a common approach is to use a $k$-balanced edge cut \cite{Huang2016}. The objective of such an approach is to minimize the proportion of edge that span partitions in a manner that balances the distribution of vertices to partitions. Intra-partition edges are referred to as \textit{local edges} and inter-partition edges are referred to as \textit{distributed edges}, [FIG]. The proportion of distributed edges is highly data dependent. However, the number of such edges is always non-negligible ranging from 15-30\%.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{./images/dist-graph}
  \caption{A partitioned graph assigned to servers in a cluster. Some edges are local and some distributed}
  \Description{Graph representation.}
  \label{dist-graph}
\end{figure}

In a distributed graph database, adjacency lists can contain ``pointers'' to edges on remote servers. This makes maintaining distributed edge reciprocal consistency more challenging - especially given a common architecture employed by contemporary distributed graph databases. Often they use an existing eventually consistent database (e.g. Apache Cassandra [REF]) to store data, which has been adapted with a programmatic API or query language expressed in terms of edges and vertices along with some gluecode to bind that interface to the underlying database. Superficially, opting for this design appears to be a good choice: the user has the modeling convenience of graphs with the operational characteristics from the underlying database. However, the problem with this design is the (lack of) transactional semantics are inherited from the underlying store. For example, Apache Cassandra provides no way of preventing updates from mutual interference in its normal multi-partition use case. Earlier work displayed how weak isolation undermines distributed edge reciprocal consistency, causing irreversible corruption that spreads at alarmingly rates \cite{Ezhilchelvan2018}.

Of course, reciprocal consistency can be maintained as a part of enforcing any of several known strong isolation guarantees (e.g., snapshot isolation) but that incurs a performance cost if users are interested only in ensuring reciprocal consistency to preserve the integrity of a distributed database. This paper addresses that challenge.

@@

A single logical write to a distributed edge consists of two sub-operations, updating reciprocal information at the source and the destination vertices. The order in which sub-operations take place are not constrained. For example, when updating an edge between A and B it is equally likely to update A then B as it is to update B then A. Without sufficient isolation, concurrent updates can interleave in a manner that violates reciprocal consistency [FIG], such an edge is known as a \emph{half-corrupted} edge. Figure [FIG] outlines the interleavings that produce half-corrupted edges.

% transactions consists of reads/writes

\subsection{\emph{Delta} Protocol}

In order to prevent all interleaving in Figure [FIG] the \emph{Delta} protocol adopts a probabilistic approach. It is assumed there is a known $\Delta$, that reflects the bound on server communication delays, with a probability $(1-\epsilon)$ that $\Delta$ is violated. Informally, the protocol consists of two-phases. In phase 1 all writes by sub-operations are temporary and a sub-operation is allowed to proceed provided there is no other temporary sub-operation preceding it within $\Delta$; measured as per the local clock time. If the attempt does not succeed, the transaction aborts, aborting any and every previous tentative write that it may have successfully completed. In phase 2, a transaction that succeeds at all its tentative write attempts, decides to commit when it completes its graph traversal; the transaction commits once all its tentative writes are committed. This protocol avoids all conflict scenarios in Figure [FIG] provided the bound is not violated. If the bound is violate corruption and its spread can still occur, Figure [FIG]. There exists an interesting relationship between $\Delta$, the proportion of aborts and the spread of corruption.

\begin{algorithm}
  \SetAlgoLined
  \eIf{sub-operation=1}{
    \eIf{$\Delta = \emptyset$}{
      Perform tentative write operation\;
      Set $\Delta$\;
      Issue sub-operation-2\;
    }{
      Abort transaction\;
    }
  }{
    \eIf{$\Delta = \emptyset$}{
      Perform tentative write operation\;
      Set $\Delta$\;
      Continue transaction\;
    }{
      Abort transaction\;
    }
  }
    \caption{\emph{Delta} Protocol Phase 1}
\end{algorithm}

@@

% Spread Corruption

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{./images/corruption_comparison}
  \caption{A partitioned graph assigned to servers in a cluster. Some edges are local and some distributed}
  \Description{Graph representation.}
  \label{dist-graph}
\end{figure}

% Aborts/s

@@

% % The bibliography is included in your source document with these two
% % commands, placed just before the \verb|\end{document}| command:
% \begin{verbatim}
% \bibliographystyle{ACM-Reference-Format}
% \bibliography{papoc}
% \end{verbatim}
% % % where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
% % % suffix, of t
%he \BibTeX\ file.

@@

% \section{Appendices}

% If your work needs an appendix, add it before the
% ``\verb|\end{document}|'' command at the conclusion of your source
% document.

% Start the appendix with the ``\verb|appendix|'' command:
% \begin{verbatim}
%   \appendix
% \end{verbatim}
% and note that in the appendix, sections are lettered, not
% numbered. This document has two appendices, demonstrating the section
% and subsection identification method.

% %%
% %% The acknowledgments section is defined using the "acks" environment
% %% (and NOT an unnumbered section). This ensures the proper
% %% identification of the section in the article metadata, and the
% %% consistent spelling of the heading.
% \begin{acks}
% To Robert, for the bagels and explaining CMYK and color spaces.
% \end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{papoc}

@@

\end{document}
\endinput
